---
title: "Project Part 1: WHO Has Water?" 
subtitle: "Kylie Sharp, Lilya Zouaoui, and Alexa Stein Repository: https://github.com/UVA-STAT3080/project-kylie-alexa-lilya"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: github_document
header-includes:
- \usepackage{setspace}
- \singlespacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyverse)
clean_water <- read.csv("stat3080_projectdata.csv", header=TRUE)
```
\section*{Introduction}
As countries develop, the assumption of their access to basic human rights should parallel. Access to water is a right, yet billions struggle to access it. The water crisis persists in society. In March 2024, 2.2 billion people lack access to safely manage water (UNICEF: 10 things you didn't know about water). Even those who do have access to water can contract diseases if the water is improperly treated. Safely managed water is crucial to a society. This issue is significant, as people are dying every day because of their lack of access to safe, clean water. Children are the most at risk with these water concerns, as every day, over 1000 children under 5 pass away from diseases linked to unsafe water, sanitation, and hygiene issues. In addition to these reasons, water is not only a necessity, but it is also a preventative measure for disease. Handwashing with soap and water is one of the simplest and most effective ways to prevent the spread of illness. Because 2.3 billion people lack handwashing capabilities due to a lack of water at home, this puts them at a disadvantage in preventing sickness. With these ideas in culmination, it cements the importance of water access to everyone. 

UNICEF notes that rural populations are the most disadvantaged in the population due to a lack of access to safe water. They cite that 8 out of 10 people who lack even basic drinking water live in rural areas ("10 things you didn't know about water"). While there has been an increase in access and a trend upward in access, the inequalities between rural and urban communities are becoming prevalent. WHO, World Bank Group, and UNICEF created a report detailing the prevalence of accelerating progress to ensure safe drinking water for all. In their explicit plan, they outline recommendations for governments, financing, capacity development, data & information, innovation, where one of their suggestions was to improve the growth of professionalized service delivery that must be supported by capacity development, particularly in small and rural systems (“State of the World’s Drinking Water”). These organizations establish the importance of the gap between the urban and rural areas. 

A UNICEF press release notes the need to continue reducing gaps in access to improved drinking water. They acknowledge that there has been an increase in drinking water and sanitation emergency issues that persist globally. In an interview with Dr. Maria Neira, a former WHO director for Public Health, Environmental and Social Determinants of Health, highlights the gaps in access depending on status: "The vast majority of those without improved sanitation are poorer people living in rural areas. Progress on rural sanitation- where it has occurred- has primarily benefited richer people, increasing inequalities.” She furthers the importance of these discrepancies in rural and urban access to water. Based on these qualified organizations, the question is whether there is a difference between these two groups. Is there a difference in the proportion of access to safely managed water (Available, Free from Contamination, and On Premises) between rural and urban settings in the sampled countries?

\section*{Author Contributions}
I developed the ggplot2 boxplot visualizations for the exploratory data analysis and authored the final results and conclusion sections, translating the output of the paired permutation test into actionable public health insights.

\section*{Data Summary}
The data represents the world, broken into countries and territories. The data attempts to summarize each country's access to safely managed water, broken into three categories: Available, Free from Contamination, and On Premises. JMP first uses estimations. These estimations start with the nationally representative data sources. In this case, the nationally representative data display the safely managed water in each country. Most of this information is collected from household interviews by national statistical offices in each of these countries. They also use new Sustainable Development Goals (SDG) indicators for safely managed drinking water, created on a facility type classification, such as whether the water comes from piped/non-piped systems. After the collection, they utilize a simple linear regression estimation, which they use to “estimate the population using different levels of service using the JMP ladder for drinking water.”

The data file created three possible values of residence: total, urban, and rural. The researcher removed the total value, as the project revolves around the urban and rural, not the combined values of these, as the total variable does not add any statistical meaning by averaging them together. In addition, the data read the safely managed elements (Available, Free from Contamination, and On Premises) as characters. To better represent the accuracy of the data, the researcher changed these value types to factors using the as.factor() function. This was necessary because these values are meant to be read as levels, where there is an order to their categorical variable, such as ascending order: Available, Free from Contamination, and On Premises. 

The data only focuses on already safely managed water, which could impact the difference between the rural and urban areas. Since the data excludes areas that do not have access to safely managed water (most likely rural and impoverished communities), it could lead to smaller differences between the two groups. To mitigate this potential issue, the researchers will emphasize that this sample is taken from countries that already have the bare minimum for safely managed water, which is the available element. By highlighting this, it will create a clear relationship between safely managed water in rural and urban areas. 

Based on the Mean Drinking Water Coverage by Residence Type and Safely Managed Element graph, there seems to be a difference in the mean coverage, depending on the residence type. The mean coverage of the available safely managed water is almost 15% lower. This indicates there may be a true difference between the two groups, possibly caused by the residence types. The research question proposes that the mean coverage is dependent on the residence type. Based on this plot's difference in values, depending on the residence type, it explores the notion of the discretion between rural and urban areas’ access to safely managed water; thus, the data appropriately answer the proposed research question.  

\section*{Exploratory Data Analysis}

```{r, echo=FALSE, warning=FALSE, message=FALSE}

water_dist <- ggplot(clean_water, aes(x= Residence.Type)) + geom_bar(fill= "lightblue", color= "black") +
  labs(
  title= "Rural and Urban Distribution",
  x= "Residence Type", 
  y= "Number of Residences", 
)

#Creating a bar graph to display the distributions of number of residence for both rural & urban areas.

clean_water_summary <- clean_water %>%
  group_by (`Residence.Type`, `Safely.managed.element`) %>%
  summarise(MeanCoverage= mean(Coverage, na.rm= TRUE), 
            .groups= "drop")

coverage_graph <- ggplot(clean_water_summary, aes(x=`Residence.Type`, y= MeanCoverage, fill= `Safely.managed.element`))+
  
geom_bar(stat= "identity", position= "dodge")+ 
  
labs(
  title= "Mean Water Coverage by Residence Type and Safely Managed Element",
  x= "Residence Type", 
  y= "Mean Coverage", 
  fill= "Safely Managed Element"
)
  
#Creating a bar graph of the mean coverage for both rural and urban areas to display the difference of distribution between the safely managed elements. 

country_avg_coverage <- clean_water %>%
  filter(`Residence.Type`%in% c("rural", "urban"))%>%
  group_by(Country)%>%
  summarise(Overall_Mean_Coverage= mean(Coverage, na.rm= TRUE), .groups = 'drop') %>% arrange (desc(Overall_Mean_Coverage))
  top_10_countries <- head(country_avg_coverage$Country, 10)
  bottom_10_countries <- tail(country_avg_coverage$Country, 10)
  
  included_countries <- c(top_10_countries, bottom_10_countries)
  
  country_data_filtered <- clean_water %>%
    filter(`Residence.Type` %in% c("rural", "urban"), Country %in% included_countries) %>% select(Country, Coverage, `Safely.managed.element`, `Residence.Type`)
  
  country_data_filtered <- country_data_filtered %>%
  left_join(country_avg_coverage, by= "Country")

coverage_graph_country <- ggplot(country_data_filtered, aes(x= reorder(Country, Overall_Mean_Coverage), y= Coverage, fill = `Residence.Type`)) + 
  
  geom_bar(stat="identity", position = "dodge")+
  coord_flip() + 
  
labs(
  title= "Water Coverage By Country",
  x= "Country",
  y= "Reported Coverage%",
  fill= "Residence Type"
)

#Creating a bar graph of the top 10 and bottom 10 countries sampled to represent the water coverage by residence type to show the inequalities between the two categories.  

bxplot1 <- ggplot(clean_water, aes(x=Safely.managed.element, y=Coverage, fill=Safely.managed.element))+ geom_boxplot(alpha=0.6) + labs(title = "Coverage Distribution by Safely Managed Element", x="Safely Managed Element", y="Coverage (%)", fill="Safely Managed Element")

#Creating a box plot of the coverage distribution by safely managed element. 


bxplot2 <- ggplot(clean_water, aes(x=Residence.Type, y=Coverage, fill=Residence.Type)) + geom_boxplot(alpha=0.6) + labs(title = "Coverage Distribution by Residence Type", x="Residence Type", y="Coverage (%)", fill="Residence Type")

#Creating a box plot of the coverage distribution by residence type. 

finalbxplot1 <- bxplot1 + geom_boxplot(alpha=0.6) + labs(title = "Coverage Distribution by Safely Managed Element", x="Safely Managed Element", y="Coverage (%)", fill="Safely Managed Element")

finalbxplot2 <- bxplot2 + geom_boxplot(alpha=0.6) + labs(title = "Coverage Distribution by Residence Type", x="Residence Type", y="Coverage (%)", fill="Residence Type")

combined_plot_graphical_1 <- (coverage_graph)

coverage_distribution <- (water_dist/coverage_graph)
coverage_distribution

combined_plot_graphical_2 <- (coverage_graph_country)
combined_plot_graphical_2

combined_plot_numerical <- (finalbxplot1 / finalbxplot2)
combined_plot_numerical


five_number <- summary(clean_water$Coverage)

five_number

```

\section*{Exploratory Data Analysis Summary}

The numerical summary of the safely managed driving water coverage shows a wide range across counties, from as low as 51% to going all the way as 100%. The median of 81% indicates that at least half the countries within this dataset have that level of coverage, but the mean being slightly lower indicates a slight left skew. This means there are a few countries that have much lower coverage, which is bringing the average down a little. From the visual graphs, the display of the top 10 and bottom 10 countries reinforces this variability. A few countries show a near-universal access, while the remainder struggle much more to provide basic levels of safe drinking water. We can see less coverage between rural areas than in urban regions through the mean coverage by residence type plot. This pattern shows that there is an unequal distribution based on residence type that favors more urban populations. The distribution plot shows that there is around an equal number of urban and rural areas, showing that the sample sizes are similar, thus appropriate to compare the two. The box plot gives us more specific data on this discrepancy between the two regions of urban and rural. The average coverage for urban areas is much higher, almost at the same level as the Q3, of the rural region. The points on the dot represent outliers in the lower end of the urban blot, whereas the fact that the rural box plot does not show that means that the lower end of the plot is not seen as outliers, as it’s a much more frequent occurrence of having lower coverage. 

The 5 number summary implies a large spread in global water coverage. The IQR being (96.62-55.97= 40.65) relatively high demonstrates the big discrepancies between the lower and higher access countries. The wide spread aligns with the bimodal tendencies seen in some of the plots, where some countries are far behind and average a similar amount. The difference in means between the urban and rural areas provides evidence of a big influence on water access. 

Contextually, these findings show that the rural-urban divide remains a big defining aspect in the global water inequality. The urban areas are benefiting from more infrastructure and a more rapid development process. This comes from a more centralized water system and consistent access that those in rural areas who depend on non-piped sources can’t access. It leaves more room for scarcity for those in rural areas. Beyond just the locality divide, there are still lots of discrepancies between water being available at a much higher rate, vs water being available and being free from contamination, which shows a much lower average. This shows much larger issues that must be discussed about water being available in some regions, but not necessarily being fully safe to consume. 

Overall the EDA indicates some clear patterns which show: Urban regions have substantially higher access to safely managed drinking water than rural regions. 
Rural populations are disproportionately disadvantaged as their count of those who live in rural areas is not much lower than those in urban areas, but they are much more at risk for not having proper water coverage. Global inequality in water access is apparent; however, there are also signs of water being accessible, but this water is not necessarily fully clean and lacks any contaminants. These show that safely managing water access is not only a global issue, but it's also very dependent on where one lives. These show the implications for the need for more targeted rural water interventions and more funding to go towards infrastructure investments in these areas, rather than just focusing on urban areas since they have more traffic. 

\section*{Methods}
To test our hypothesis we used a paired permutation test, where: Null Hypothesis: The true proportion of access to safely managed water in rural settings is equal to the proportion in urban settings. Alternative Hypothesis: The proportion of access to safely managed water in rural areas is not equal to the proportion in urban settings.

A permutation test is a non-parametric, assumption light statistical test that assesses the null hypothesis by repeatedly testing and shuffling the values of data between groups to create a null distribution.The use of the permutation test is relevant for a two-sample data set that is dependent. The use of a non-parametric over a parametric is used since it does not assume the underlying distribution of the proportions is normal. In the data, many countries display a range of extreme values ranging between 0 and 1. The normal approximations required by classical approaches are often violated in cases like that, so to avoid violating assumptions, the permutation is a better choice, as it will provide statistically sound inferences for the difference in proportions. Within the permutation test, the researchers are sampling rural and urban areas, meaning that the samples are paired; thus, the use of the paired permutation test is appropriate.Since we are using permutation tests which are non-parametric we don't need to worry about verifying the distributional or sample-size assumptions.

The first assumption is independence. We need to check if our pairs (countries) are independent to ensure there are no repeated measurements and that the countries don't influence each other. Since each country only appears once in the dataset, the assumption for independence is met. Next we look at the Exchangeability assumption under the null needs to be met as the next assumption as stated above. If rural and urban access were truly equal, the sign of each paired difference would be random. Since we see that both measurements come from the same country and are directly comparable the sign-flipping becomes valid. The third assumption is that they have to have an equal measurement scale. Since both sets of data for rural and urban estimates use the same WHO/UNICEF methodology we can compare the two under equal scales. 

\section*{Results}
```{r, echo = FALSE, warning=FALSE, message=FALSE}
clean_water_prop <- clean_water |>
  
# Grouping the data by country and the type of residence
  group_by(ISO3, Country, `Residence.Type`) |>
  summarise(
# Calculating the mean of the coverage column for each country-residence type combination
    Overall_Prop = mean(Coverage),
    .groups = 'drop'
) |>
# Using residence.type to create separate columns, filling the new columns with the calculated proportions, and adding a prefix to the new columns
  pivot_wider(
    names_from ='Residence.Type',
    values_from = Overall_Prop,
    names_prefix = "Prop_"
) |>
  rename(
  Rural_Prop = Prop_rural,
  Urban_Prop = Prop_urban
) |>
#The proportion difference is the data used in the paired test
  mutate(
    Prop_difference = Urban_Prop- Rural_Prop
)

# Isolate the column of observed differences, remove NA values, and determine the sample size 
difference <- clean_water_prop$Prop_difference
difference <- difference[!is.na(difference)]
n <- length(difference)
# Calculate test statistic (observed mean difference)
observed_mean_diff <- mean(difference)
# Set seed and choose number of permutations
set.seed(51)
K <- 100000
# Generate a null distribution by repeatedly randomizing the sign of each paired difference (this simulates the null hypothesis where the sign of each urban-rural difference is equally likely to be positive or negative)
null_dist <- replicate(K,{
  signs <- sample(c(-1,1), n, replace = TRUE)
  mean(difference*signs)
})
# Calculate the p-value (the proportion of null means at least as extreme as the observed)
p_val <- mean(abs(null_dist) >= abs(observed_mean_diff))
p_val
observed_mean_diff
```

\section*{Conclusions}

The test statistic (the observed mean difference in coverage) is about 13.05, which means that, on average, urban areas have 13.05 percentage points higher clean-water coverage than rural areas. None of the permuted differences were as extreme as the observed value, the p-value result was 0, meaning there is statistically significant evidence that urban and rural clean-water coverage differ. This is not due to random sampling variability since with a large sample of permutations we would not see a difference larger than the observed p value. Overall, the test was statistically significant and it provides enough statistical evidence to show that the proportion of populations with access to safely managed water differs between rural and urban settings. Our results reinforce WHO and UNICEF claims that people living in rural areas systematically experience worse access to safely managed water (available on premises and free from contamination). A gap of about 13 percentage points is large: even when national coverage improves, those gains are not distributed evenly, and rural communities are left behind, with further issues for health, hygiene, and economic opportunity occurring later on.

One limitation is our data is that we are analyzing a summary values per country. This means our analysis is not accounting for the individual variation within each country across various regions or income groups which may alter the definitions between our rural and urban areas. Another limitation is that the WHO water-access estimates rely model estimates based on the data quality and standards for each country, however these standards may differ across each country or differ in how precise the estimates are in different regions. Lastly, our countries are all treated as individual variables, however sometimes neighboring countries may share infrastructure or certain political standards that may lead to possible correlations. 

Our analysis still offers a level of evidence that is consistent with global reports. Even if there may be some variability in standards, the majority of the countries follow a standardized WHO and UNICEF definitions within the data. This means that the pattern we see in our results can be generalized beyond our specific sample even if the exact size of the gap varies. Generalizing from a policy perspective is noting that simply rating national averages is not sufficient. We need to see more targeted investments in more rural water infrastructure and delivery to actually see more progress in equitable water coverage rather than just increasing disparities across regions.  



\newpage


\section*{References}
10 things you didn’t know about water. UNICEF. (2024, March 14). https://www.unicef.org/stories/10-things-you-didnt-know-about-water

Methods. Methods | JMP. (n.d.). https://washdata.org/topics/methods 

OpenAI. (2023). ChatGPT (Mar 14 version) [Large language model]. https://chat.openai.com/chat

State of the World’s Drinking Water. World health Organization, World Bank Group, Unicef. (n.d.). https://cdn.who.int/media/docs/default-source/wash-documents/water-safety-and-quality/state-of-drinking-water-report_ex-summary_english.pdf?sfvrsn=9cb7ca6d_4&download=true 

Water Access by Safely Managed Element by Country. JMP. (n.d.). https://washdata.org/data/household#!/table?geo0=region&geo1=sdg 


WHO/UNICEF highlight need to further reduce gaps in access to improved drinking water and Sanitation. UNICEF. (2014, May 8). https://www.unicef.org/turkiye/en/press-releases/whounicef-highlight-need-further-reduce-gaps-access-improved-drinking-water-and